Solving the maximum independant set problem by generic programming einsum networks

### 2021-06-12 paper/paper.tex
update code

Before contracting the einsum network and evaluating the independence polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A semiring is a ring without additive inverse, while a commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relation must hold for arbituary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} * a = \mymathbb{0}
\end{align*}

@@

@@

\section{Technical guide}
\begin{description}
	\item[OMEinsum] a package for einsum,
	\item[OMEinsumContractionOrders] a package for finding the optimal contraction order for einsum \\ \href{https://github.com/Happy-Diode/OMEinsumContractionOrders.jl}{https://github.com/Happy-Diode/OMEinsumContractionOrders.jl},
	\item[TropicalGEMM] a package for efficient tropical matrix multiplication (compatible with OMEinsum),
	\item[TropicalNumbers] a package providing tropical number types and tropical algebra, one o the dependency of TropicalGEMM,
	\item[LightGraphs] a package providing graph utilities, like random regular graph generator,
	\item[Polynomials] a package providing polynomial algebra and polynomial fitting,
	\item[Mods and Primes] packages providing finite field algebra and prime number generators.
\end{description}

One can install these packages by opening a julia REPL, type \colorbox{lightgray}{\texttt{]}} to enter the \texttt{pkg>} mode and type, e.g.
\begin{lstlisting}
pkg> add OMEinsum
\end{lstlisting}

An example of computing the maximum independant set size
\begin{lstlisting}
julia> using OMEinsum, LightGraphs, TropicalNumbers

julia> n, k = 6, 3
(6, 3)

julia> g = LightGraphs.random_regular_graph(n, k)
{6, 9} undirected simple Int64 graph

julia> ixs = [minmax(e.src,e.dst) for e in LightGraphs.edges(g)]
9-element Vector{Tuple{Int64, Int64}}:
 (1, 2)
 (1, 4)
 (1, 5)
 (2, 3)
 (2, 6)
 (3, 5)
 (3, 6)
 (4, 5)
 (4, 6)

julia> code = EinCode((ixs..., [(i,) for i in LightGraphs.vertices(g)]...), ())
1∘2, 1∘4, 1∘5, 2∘3, 2∘6, 3∘5, 3∘6, 4∘5, 4∘6, 1, 2, 3, 4, 5, 6 ->

julia> optimized_code = optimize_greedy(code, Dict([i=>2 for i=1:6]))
4∘6, 4∘6 -> 
├─ 4∘6, 6 -> 4∘6
│  ├─ 6
│  └─ 4∘6
└─ 2∘5∘4, 5∘2∘6 -> 4∘6
   ├─ 2∘5∘6, 6∘2 -> 5∘2∘6
   │  ├─ 2∘6, 2 -> 6∘2
   │  │  ├─ 2
   │  │  └─ 2∘6
   │  └─ 2∘3, 5∘6∘3 -> 2∘5∘6
   │     ├─ 3∘5, 6∘3 -> 5∘6∘3
   │     │  ├─ 3∘6, 3 -> 6∘3
   │     │  │  ⋮
   │     │  │  
   │     │  └─ 3∘5, 5 -> 3∘5
   │     │     ⋮
   │     │     
   │     └─ 2∘3
   └─ 2∘4∘1, 1∘4∘5 -> 2∘5∘4
      ├─ 1∘5, 5∘4 -> 1∘4∘5
      │  ├─ 4∘5, 4 -> 5∘4
      │  │  ├─ 4
      │  │  └─ 4∘5
      │  └─ 1∘5
      └─ 2∘1, 1∘4 -> 2∘4∘1
         ├─ 1∘4
         └─ 1∘2, 1 -> 2∘1
            ├─ 1
            └─ 1∘2

julia> function mis_contract(code::OMEinsum.NestedEinsum, x::T) where T
           tensors = map(OMEinsum.getixs(Iterators.flatten(code))) do ix
               @assert length(ix) == 1 || length(ix) == 2
                           length(ix) == 1 ? [one(T), x] : [one(T) one(T); one(T) zero(T)]
           end
           code(tensors...)
       end
mis_contract (generic function with 1 method)

julia> result = mis_contract(optimized_code, Tropical(1.0))
0-dimensional Array{TropicalF64, 0}:
2.0ₜ
\end{lstlisting}

For larger graph, we hight recommend using two additional packages,
\texttt{TropicalGEMM} for BLAS speed tropical matrix multiplication and
\texttt{OMEinsumContractionOrders} for hyper optimized contraction order (the KaHyPar~\cite{Schlag2020} approach) ~\cite{Kourtis2019,Gray2021,Pan2021}.

\section{Indenpendence polynomials}
\subsection{Common}
\begin{lstlisting}
using Polynomials

_auto_mispolytensor(x::T, ix::NTuple{2}) where T = T[1 1; 1 0]
_auto_mispolytensor(x::T, ix::NTuple{1}) where T = T[1, x]
function generate_polyxs!(x::T, code::OMEinsum.NestedEinsum, xs=Vector{Any}(undef, ninput(code))) where {T}
    for (ix, arg) in zip(OMEinsum.getixs(code.eins), code.args)
		if arg isa Integer
			xs[arg] = _auto_mispolytensor(x, ix)
		else
        	generate_polyxs!(x, arg, xs)
		end
    end
	return xs
end

function generate_polyxs!(x::T, code::EinCode, xs=Vector{Any}(undef, ninput(code))) where {T}
    for (i,ix) in enumerate(OMEinsum.getixs(code))
		xs[i] = _auto_mispolytensor(x, ix)
    end
	return xs
end

function mis_polysolve(code, x::T) where {T}
	xs = generate_polyxs!(x, code, Vector{Any}(undef, NoteOnTropicalMIS.ninput(code)))
	code(xs...)
end
\end{lstlisting}

\subsection{Symbolic}
\begin{lstlisting}
function independence_polynomial(::Val{:polynomial}, code)
    mis_polysolve(code, Polynomial([0, 1.0]))[]
end
\end{lstlisting}

\subsection{Fitting and FFT}
\begin{lstlisting}
using FFTW
function independence_polynomial(::Val{:fft}, code; mis_size=Int(mis_solve(code)[].n), r=1.0)
	ω = exp(-2im*π/(mis_size+1))
	xs = r .* collect(ω .^ (0:mis_size))
	ys = [mis_polysolve(code, x)[] for x in xs]
	Polynomial(real.(ifft(ys) ./ (r .^ (0:mis_size))))
end

function independence_polynomial(::Val{:fitting}, code; mis_size=Int(mis_solve(code)[].n))
	xs = (0:mis_size)
	ys = [mis_polysolve(code, x)[] for x in xs]
	fit(xs, ys, mis_size)
end
\end{lstlisting}

\subsection{Finite field algebra}
\begin{lstlisting}
using Mods, Primes

# pirate
Base.abs(x::Mod) = x
Base.isless(x::Mod{N}, y::Mod{N}) where N = mod(x.val, N) < mod(y.val, N)

function _independance_polynomial(::Type{T}, code, mis_size::Int) where T
	xs = 0:mis_size
	ys = [mis_polysolve(code, T(x))[] for x in xs]
	A = zeros(T, mis_size+1, mis_size+1)
	for j=1:mis_size+1, i=1:mis_size+1
		A[j,i] = T(xs[j])^(i-1)
	end
	A \ T.(ys)
end

function independence_polynomial(::Val{:finitefield}, code; mis_size=Int(mis_solve(code)[].n), max_order=100)
    N = typemax(Int)
    YS = []
    local res
    for k = 1:max_order
	    N = prevprime(N-1)
        T = Mods.Mod{N,Int}
        rk = _independance_polynomial(T, code, mis_size)
        push!(YS, rk)
        if max_order==1
            return Polynomial(Mods.value.(YS[1]))
        elseif k != 1
            ra = improved_counting(YS[1:end-1])
            res = improved_counting(YS)
            ra == res && return Polynomial(res)
        end
    end
    @warn "result is potentially inconsistent."
    return Polynomial(res)
end

function improved_counting(sequences)
    map(yi->Mods.CRT(yi...), zip(sequences...))
end
\end{lstlisting}

\section{Maximum independent set}
\subsection{Tropical numbers}
\begin{lstlisting}
function independence_polynomial(::Val{:polynomial}, code)
    mis_polysolve(code, Polynomial([0, 1.0]))[]
end
\end{lstlisting}

\subsection{Enumerating configurations}

The following is a complete example.

\begin{lstlisting}
using OMEinsum, TropicalNumbers
using OMEinsum: NestedEinsum

struct ConfigEnumerator{N,C}
    data::Vector{StaticBitVector{N,C}}
end

Base.length(x::ConfigEnumerator{N}) where N = length(x.data)
Base.:(==)(x::ConfigEnumerator{N,C}, y::ConfigEnumerator{N,C}) where {N,C} = x.data == y.data

function Base.:+(x::ConfigEnumerator{N,C}, y::ConfigEnumerator{N,C}) where {N,C}
    res = ConfigEnumerator{N,C}(vcat(x.data, y.data))
    return res
end

function Base.:*(x::ConfigEnumerator{L,C}, y::ConfigEnumerator{L,C}) where {L,C}
    M, N = length(x), length(y)
    z = Vector{StaticBitVector{L,C}}(undef, M*N)
    for j=1:N, i=1:M
        z[(j-1)*M+i] = x.data[i] .| y.data[j]
    end
    return ConfigEnumerator{L,C}(z)
end

Base.zero(::Type{ConfigEnumerator{N,C}}) where {N,C} = ConfigEnumerator{N,C}(StaticBitVector{N,C}[])
Base.one(::Type{ConfigEnumerator{N,C}}) where {N,C} = ConfigEnumerator{N,C}([TropicalNumbers.staticfalses(StaticBitVector{N,C})])

function enumerator_t(::Type{T}, ix::NTuple{2}, vertex_index) where {T1, N, C, T<:CountingTropical{T1,ConfigEnumerator{N,C}}}
    [one(T) one(T); one(T) zero(T)]
end
function enumerator_t(::Type{T}, ix::NTuple{1}, vertex_index) where {T1, N, C, T<:CountingTropical{T1,ConfigEnumerator{N,C}}}
    [one(T), CountingTropical(one(T1), ConfigEnumerator([TropicalNumbers.onehot(StaticBitVector{N,C}, vertex_index[ix[1]])]))]
end
function enumerator_t(::Type{T}, ix::NTuple{2}, vertex_index) where {T1,N,C, T<:ConfigTropical{T1,N,C}}
    [one(T) one(T); one(T) zero(T)]
end
function enumerator_t(::Type{T}, ix::NTuple{1}, vertex_index) where {T1,N,C, T<:ConfigTropical{T1,N,C}}
    [one(T), T(one(T1), TropicalNumbers.onehot(StaticBitVector{N,C}, vertex_index[ix[1]]))]
end

function generate_xs!(f, x::T, code::NestedEinsum, xs) where {T}
    for (ix, arg) in zip(OMEinsum.getixs(code.eins), code.args)
		if arg isa Integer
			xs[arg] = f(x, ix)
		else
        	generate_xs!(f, x, arg, xs)
		end
    end
	return xs
end

function generate_xs!(f, x::T, code::EinCode, xs) where {T}
    for (i,ix) in enumerate(OMEinsum.getixs(code))
		xs[i] = f(x, ix)
    end
	return xs
end

symbols(::EinCode{ixs}) where ixs = unique(Iterators.flatten(ixs))
symbols(ne::NestedEinsum) = symbols(Iterators.flatten(ne))
ninput(::EinCode{ixs}) where ixs = length(ixs)
ninput(ne::NestedEinsum) = ninput(Iterators.flatten(ne))
function mis_config(code; all=false)
    vertex_index = Dict([s=>i for (i, s) in enumerate(symbols(code))])
    N = length(vertex_index)
    C = TropicalNumbers._nints(N)
    T = all ? CountingTropical{Float64, ConfigEnumerator{N,C}} : ConfigTropical{Float64, N, C}
	xs = generate_xs!((T, ix)->enumerator_t(T, ix, vertex_index), T, code, Vector{Any}(undef, ninput(code)))
	code(xs...)
end

mis_config(optimize_greedy(ein"i,j,k,l,ij,jk,kl,li->", Dict('i'=>2, 'j'=>2, 'k'=>2, 'l'=>2)); all=true)  # square
\end{lstlisting}

You will see the following output

\begin{lstlisting}
0-dimensional Array{CountingTropical{Float64, ConfigEnumerator{4, 1}}, 0}:
(2.0, ConfigEnumerator{4, 1}(StaticBitVector{4, 1}[1010, 0101]))ₜ
\end{lstlisting}

\end{document}

>>>>

Before contracting the einsum network and evaluating the independence polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A semiring is a ring without additive inverse, while a commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relation must hold for arbituary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}

@@

\section{discussion}
We introduced in the main text how to compute the indenpendence polynomial, maximum independent set and optimal configurations.
It is interesting that although these properties are global,
they can be solved by designing different element types that having two operations $\oplus$ and $\odot$ and two special elements $\mymathbb{0}$ and $\mymathbb{1}$.
One thing in common is that they all defines a commutative semiring.
Here, we want the $\oplus$ and $\odot$ operations being commutative because we do not want the contraction result of an einsum network to be sensitive to the contraction order.
We show most of the implementation in Appendix \ref{sec:technical}. It is supprisingly short.
The style that we program is called generic programming,
it is about writing a single copy of code, feeding different types into it, and the program computing the result with a proper performance.
It is language dependent feature. If someone want to implement this algorithm in python,
one has to rewrite the matrix multiplication for different element types in C and then export the interface to python.
In C++, users can use templates for such a purpose.
In our work, we chose Julia because its just in time compiling is very powerful that it can generate fast code dynamically for users.
Elements of fixed size, such as the finite field algebra, tropical number, tropical number with counting/configuration field used in the main text can be inlined in an array.
Furthermore, these inlined arrays can be upload to GPU devices for faster generic matrix multiplication implemented in CUDA.jl.

@@

\section{Technical guide}\label{sec:technical}
\begin{description}
	\item[OMEinsum] a package for einsum,
	\item[OMEinsumContractionOrders] a package for finding the optimal contraction order for einsum \\ \href{https://github.com/Happy-Diode/OMEinsumContractionOrders.jl}{https://github.com/Happy-Diode/OMEinsumContractionOrders.jl},
	\item[TropicalGEMM] a package for efficient tropical matrix multiplication (compatible with OMEinsum),
	\item[TropicalNumbers] a package providing tropical number types and tropical algebra, one o the dependency of TropicalGEMM,
	\item[LightGraphs] a package providing graph utilities, like random regular graph generator,
	\item[Polynomials] a package providing polynomial algebra and polynomial fitting,
	\item[Mods and Primes] packages providing finite field algebra and prime number generators.
\end{description}

One can install these packages by opening a julia REPL, type \colorbox{lightgray}{\texttt{]}} to enter the \texttt{pkg>} mode and type, e.g.
\begin{lstlisting}
pkg> add OMEinsum LightGraphs Mods Primes FFTW Polynomials TropicalNumbers
\end{lstlisting}

It may supprise you that the Julia implementation of algorithms introduced in the paper is so short that except the bounding and sparsity related parts,
all are contained in this appendix. After installing required packages, one can open a Julia REPL and copy the following code into it.

\lstinputlisting[breaklines]{../democode/demo.jl}

In the above examples, the configuration enumeration is very slow, one should use the optimal MIS size for bounding as decribed in the main text.
We will not show any example about implementing the backward rule here because it has approximately 100 lines of code.
Please checkout our Github repository \href{https://github.com/Happy-Diode/NoteOnTropicalMIS}{https://github.com/Happy-Diode/NoteOnTropicalMIS}.
\end{document}
